Updated Constraints on Self-Interacting Dark Matter from Supernova 1987A

### 2017-01-05 draft.tex
Added two paragraphs to the introduction.

\documentclass[nofootinbib,prd,superscriptaddress,twocolumn]{revtex4}
\usepackage{amsmath,graphicx,hyperref}
\usepackage{slashed}
\usepackage{epstopdf}

@@

@@

@@

Dark matter that interacts with itself through a large elastic scattering cross section 
(on the order of $\sigma \sim 10^{-24}\, \mathrm{cm}^2$) 
has been considered for numerous reasons over the past several decades.

This calculation was done in Ref.~\cite{Dent:2012mx} Using the one-pion exchange model of nucleon-nucleon interactions\footnote{There is also some question as to the applicability of this approach in this regime. Pionless effective field theory techniques developed for low-energy nucleons might provide a more accurate answer; this may be considered in a later paper} they largely followed the form of the calculation of the axionic bremsstrahlung rate [brinkmann/turner/iwamoto], in which the axion is taken to be massless. Considering the energy scales in the problem, this is clearly a reasonable approximation when dealing with axion masses on the order of $ 10^-3 $ eV; it is less clear that it is still appropriate when dealing with dark photon masses on the order of 10-100 MeV. We therefore reproduce the calculation without making that simplifying assumption, and see if it has an effect on the result. We find that it does indeed quite dramatically change the excluded region in $ \eps-m_A $ space, increasing the lower bound by some two orders of magnitude and opening up a considerable portion of the parameter space.

@@

The calculation of these diagrams is straightforward, and is described in the appendix. Having obtained the diagrams the calculation diverges from the axion bremsstrahlung calculation, since in the kinematic region of interest for supernova explosions the mass of the $A'$ boson is not necessarily negligible. The kinematic relations are thus \bea 
p_1 \cdot p_2 &=& M_N^2 - \frac{l^2}{2} - \frac{k^2}{2} + p_2 \cdot q_a,\\
p_1 \cdot p_3 &=& M_N^2 + k \cdot l - \frac{k^2}{2} + p_3 \cdot q_a,\\  
p_1 \cdot p_4 &=& k \cdot l + M_N^2 - \frac{l^2}{2} + p_4 \cdot q_a, \\
p_2 \cdot p_3 &=& M_N^2 - \frac{l^2}{2}, \\ 
p_2 \cdot p_4 &=& M_N^2 - \frac{k^2}{2},\\
p_3 \cdot p_4 &=& k \cdot l + M_N^2 - \frac{l^2 + k^2}{2}.
\eea
Unfortunately, with these relations it does not seem to be possible to simplify the amplitude into an analytically usable form without making unjustified approximations, and so the final expression for the spin-averaged squared matrix element, $ \mathcal{M}^2_{pp}$, contains some two hundred and fifty terms, and thus we do not reproduce that result here. The pn calculation likewise results in a very large and unwieldy expression for $ \mathcal{M}^2_{pn}$. \footnote{At this point we diverge from Ref.~\cite{Dent:2012mx} for two reasons.  One, they use the kinematics as if the $A'$ were massless, as in the axion calculation, and independent of that choice, their squared matrix element does not respect the correct exchange symmetry, as can be seen by inspection of their Eq. (A.3), with their claim that only the coefficient $C_k$ appearing in that equation is nonzero.}

@@

\subsection{Trapping limit}
The second constraint that produces an upper bound on the excluded region comes from considering trapping of dark bosons within the supernova. With a large enough coupling, the new particles will thermalize and then will be emitted from a spherical shell.  In this case the luminosity is given simply by the Steffan-Boltzmann law
\beq
\mathcal{L}_t  = 4\pi r^2 T_A^4 \sigma,
\eeq
where $ r $ is now the radius of the emitting shell and $ T_A $ its temperature. We can estimate $ r = 10$ km, since the density of the supernova drops drastically around that point. After taking that value for r, the bound on the luminosity translates into a bound on $ T_A $ 
\beq
T_A \le 9.586\rm\ MeV.
\eeq
That bound can then be translated into the desired bound on the coupling as a function of mass by assuming that the particles are emitted from an optical depth $ \tau = 2/3 $, and finding the temperature that corresponds to that optical depth. This is a somewhat involved calculation.
First, one needs a model for the density and temperature in the supernova. Following \cite{Dent:2012mx}, we assume 
\bea
\rho &=& \rho_p \left(\frac{R}{r}\right)^n,\\
T &=& T_R \left(\frac{\rho(r)}{\rho_R}\right)^{1/3},
\eea
with $  \rho_p  = 3\times10^{14}\ {\rm g/cm}^3, T_R = 30$ MeV, and taking $ n = 5 $. The optical depth is given by 
\beq
\tau = \int_{r_x}^{\infty} \kappa \rho dr,
\eeq
where $ \kappa$ is the opacity.  To find the opacity, we start from the reduced mean Rosseland opacity 
\beq
\frac{1}{\kappa \rho} = \int_{m_x}^{\infty} \frac{15}{4 \pi^4 T^5} \frac{E_A^2 e^{E_A/T} \sqrt{E_A^2 - m_A^2}}{(e^{E_A/T}-1)^2} l_A dE_A,
\eeq
where $ l_A $ is the mean free path.

@@

\begin{figure}[h!]
\includegraphics[height=9cm]{stages.eps}
\caption{The changes in the bounds generated as each stage of the calculation departs from Ref.~\cite{Dent:2012mx}. Interestingly the kinematics have apparently relatively little effect.}
\end{figure}

\begin{figure}[h!]
	\includegraphics[height=9cm]{endtoend.png}
	\caption{A comparison of the final result for the excluded region with the result from Ref.~\cite{Dent:2012mx}. The excluded region is clearly significantly less restrictive, and opens up the region of parameter space for $\epsilon \approx  1e-8-9 $.}
\end{figure}

@@

\appendix
\section{Bremsstrahlung diagrams}
	Each of the $p-p-\pi $ vertices contribute a factor of $g_{pp}$,  the pseudoscalar pion-nucleon coupling. $f_{pp}$, the pseudovector pion-nucleon coupling, is related by $g_{pp} = \frac{2M_N}{m_\pi} f_{pp} $. The $ p-p-A' $ vertex contributes $e\epsilon \gamma_\mu$, where $\epsilon$ is the mixing strength between the photon and dark photon. Labeling the contributions according to the Feynman diagrams in Fig. , the matrix elements are 
	\bea
	M_1 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_1} \bar{u}(p_3) \gamma_5 u(p_2) \bar{u}(p4) \gamma_5 (\slashed{p}_1 - q_a +M_N)\slashed{\varepsilon} u(p_1),\\
	M_2 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_1} \bar{u}(p_4) \gamma_5 u(p_2) \bar{u}(p3) \gamma_5 (\slashed{p}_1 - q_a +M_N)\slashed \varepsilon  u(p_1), \\
	M_3 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_2} \bar{u}(p_3) \gamma_5 u(p_1) \bar{u}(p4) \gamma_5 (\slashed{p}_2 - q_a +M_N)\slashed \varepsilon  u(p_2), \\
	M_4 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_2} \bar{u}(p_4) \gamma_5 u(p_1) \bar{u}(p3) \gamma_5 (\slashed{p}_2 - q_a +M_N)\slashed \varepsilon  u(p_2), \\
	M_5 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_3} \bar{u}(p_4) \gamma_5 u(p_1) \bar{u}(p3) \slashed{\varepsilon} (\slashed{p}_3 + q_a +M_N)\gamma_5 u(p_2),\\
	M_6 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_3} \bar{u}(p_3) \gamma_5 u(p_1) \bar{u}(p4) \slashed{\varepsilon} (\slashed{p}_3 + q_a +M_N)\gamma_5 u(p_2),\\
	M_7 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_4} \bar{u}(p_3) \gamma_5 u(p_2) \bar{u}(p4) \slashed{\varepsilon} (\slashed{p}_4 + q_a +M_N)\gamma_5 u(p_1),\\
	M_8 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_4} \bar{u}(p_4) \gamma_5 u(p_2) \bar{u}(p_3) \slashed{\varepsilon} (\slashed{p}_4 + q_a +M_N)\gamma_5 u(p_1),
	\eea
	where the exchange momenta are defined by $ k = p_2-p_4 $ and $ l = p_2-p_3 $, and the dark photon's polarization is given by $\varepsilon$. 
	For $pn$ scattering, four of the diagrams are simply the same as those given above. The only new Feynman diagram comes from bremsstrahlung off the internal pion, and results in  
	 \beq
	 M'_5 = \frac{4 M_N}{ m_\pi} \frac{f_{pn}^2 e \epsilon}{l^2-m_\pi^2
	 }  \frac{1}{(l-q_a)^2 - m_\pi^2} \bar{u}(p_4) \gamma_5 u(p_1) \bar{u}(p_3) u(p_2) (q_a - 2l)\cdot \varepsilon
	 \eeq
	 Note that these matrix elements agree with Ref.~\cite{Dent:2012mx}.
\acknowledgments

@@

\begin{thebibliography}{99}
%\normalsize

%\cite{Dent:2012mx}
\bibitem{Dent:2012mx} 
  J.~B.~Dent, F.~Ferrer and L.~M.~Krauss,
  %``Constraints on Light Hidden Sector Gauge Bosons from Supernova Cooling,''
  arXiv:1201.2683 [astro-ph.CO].
  %%CITATION = ARXIV:1201.2683;%%
  %56 citations counted in INSPIRE as of 29 Jul 2016

\end{thebibliography}

>>>>

\documentclass[nofootinbib,prd,superscriptaddress,twocolumn]{revtex4}
\usepackage{amsmath,graphicx,hyperref}
\usepackage{slashed}
\usepackage{epstopdf}
\usepackage{color}
\usepackage[dvipsnames]{xcolor}

@@

%--- journals
\newcommand{\physrep}{Phys. Reports}
\newcommand{\mnras}{Mon. Not. R. Astron. Soc.}
\newcommand{\apjl}{Astrophys. J. Lett.}
\newcommand{\jcap}{J. Cosmol. Astropart. Phys.}

%--- for comments
\newcommand{\arz}[1]{{{\bf{\color{BrickRed}[ARZ: #1]}}}}

@@

\bibliographystyle{apsrev}

@@

A nearly overwhelming preponderance of observational evidence indicates that a form of nonrelativistic, nonbaryonic, 
dark matter constitutes the vast majority of mass in the Universe and drives the formation of cosmic structure. 
The pace of the quest to identify the dark matter is accelerating on many fronts. Weakly-interacting massive 
particles (WIMPs) have received the most attention as dark matter candidates (see Ref.~\cite{jungman_etal96} for a review). 
Dark matter particles that interact with standard model particles only weakly, while interacting among themselves 
much more strongly have been studied as an alternative to WIMP scenarios in many contexts 
\cite{carlson_etal92,deLaix_etal95,atrio-barandela_davidson97,spergel_steinhardt00,hogan_dalcanton00,mohapatra_teplitz00,
dave_etal01,hisano_etal04,hisano_etal05,pospelov_etal08,arkani-hamed_etal08a,lattanzi_silk08,ackerman_etal09,feng_etal09,
kong_etal15} \arz{Need to add some more recent references here.} 
and constraints on self-interacting dark matter (SIDM) models have been explored by many authors \cite{yoshida_etal00,gnedin_ostriker01,miralda-escude02,randall_etal08,kamionkowski_profumo08,zentner09,robertson_zentner09,pieri_etal09,spolyar_etal09,finkbeiner_etal09,
slatyer_etal09,bramante_etal14,albuquerque_etal14,kaplinghat_etal14,chen_etal14,feng_etal16,catena_widmark16}.
In this paper, we revisit and update astrophysical constraints on SIDM models from 
supernova cooling, finding constraints that are considerably less restrictive than in 
previous work. 
\arz{Might want to add a comment about SIDM affecting small-scale structure in CDM, the satellites problems and core/cusp problems.}

SIDM models in which large self-interaction cross sections are mediated by sufficiently light 
bosons ($M \lesssim 100$~GeV) can be constrained astrophysically using supernovae, particularly 
SN1987A. Light gauge bosons will be produced within the hot supernovae core and 
radiated from the supernova. This non-standard energy loss mechanism can result in an energy loss rate 
from the supernova core that is inconsistent with observations of SN1987A, analogous to the 
classic constraint on axions \cite{raffelt96_book}. Constraints on SIDM from this mechanism 
were already considered by \cite{dent_etal12} and \cite{rrapaj_reddy16}.

This calculation was done in Ref.~\cite{dent_etal12} Using the one-pion exchange model of nucleon-nucleon interactions\footnote{There is also some question as to the applicability of this approach in this regime. Pionless effective field theory techniques developed for low-energy nucleons might provide a more accurate answer; this may be considered in a later paper} they largely followed the form of the calculation of the axionic bremsstrahlung rate [brinkmann/turner/iwamoto], in which the axion is taken to be massless. Considering the energy scales in the problem, this is clearly a reasonable approximation when dealing with axion masses on the order of $ 10^-3 $ eV; it is less clear that it is still appropriate when dealing with dark photon masses on the order of 10-100 MeV. We therefore reproduce the calculation without making that simplifying assumption, and see if it has an effect on the result. We find that it does indeed quite dramatically change the excluded region in $ \eps-m_A $ space, increasing the lower bound by some two orders of magnitude and opening up a considerable portion of the parameter space.

@@

The calculation of these diagrams is straightforward, and is described in the appendix. Having obtained the diagrams the calculation diverges from the axion bremsstrahlung calculation, since in the kinematic region of interest for supernova explosions the mass of the $A'$ boson is not necessarily negligible. The kinematic relations are thus \bea 
p_1 \cdot p_2 &=& M_N^2 - \frac{l^2}{2} - \frac{k^2}{2} + p_2 \cdot q_a,\\
p_1 \cdot p_3 &=& M_N^2 + k \cdot l - \frac{k^2}{2} + p_3 \cdot q_a,\\  
p_1 \cdot p_4 &=& k \cdot l + M_N^2 - \frac{l^2}{2} + p_4 \cdot q_a, \\
p_2 \cdot p_3 &=& M_N^2 - \frac{l^2}{2}, \\ 
p_2 \cdot p_4 &=& M_N^2 - \frac{k^2}{2},\\
p_3 \cdot p_4 &=& k \cdot l + M_N^2 - \frac{l^2 + k^2}{2}.
\eea
Unfortunately, with these relations it does not seem to be possible to simplify the amplitude into an analytically usable form without making unjustified approximations, and so the final expression for the spin-averaged squared matrix element, $ \mathcal{M}^2_{pp}$, contains some two hundred and fifty terms, and thus we do not reproduce that result here. The pn calculation likewise results in a very large and unwieldy expression for $ \mathcal{M}^2_{pn}$. \footnote{At this point we diverge from Ref.~\cite{dent_etal12} for two reasons.  One, they use the kinematics as if the $A'$ were massless, as in the axion calculation, and independent of that choice, their squared matrix element does not respect the correct exchange symmetry, as can be seen by inspection of their Eq. (A.3), with their claim that only the coefficient $C_k$ appearing in that equation is nonzero.}

@@

\subsection{Trapping limit}
The second constraint that produces an upper bound on the excluded region comes from considering trapping of dark bosons within the supernova. With a large enough coupling, the new particles will thermalize and then will be emitted from a spherical shell.  In this case the luminosity is given simply by the Steffan-Boltzmann law
\beq
\mathcal{L}_t  = 4\pi r^2 T_A^4 \sigma,
\eeq
where $ r $ is now the radius of the emitting shell and $ T_A $ its temperature. We can estimate $ r = 10$ km, since the density of the supernova drops drastically around that point. After taking that value for r, the bound on the luminosity translates into a bound on $ T_A $ 
\beq
T_A \le 9.586\rm\ MeV.
\eeq
That bound can then be translated into the desired bound on the coupling as a function of mass by assuming that the particles are emitted from an optical depth $ \tau = 2/3 $, and finding the temperature that corresponds to that optical depth. This is a somewhat involved calculation.
First, one needs a model for the density and temperature in the supernova. Following \cite{dent_etal12}, we assume 
\bea
\rho &=& \rho_p \left(\frac{R}{r}\right)^n,\\
T &=& T_R \left(\frac{\rho(r)}{\rho_R}\right)^{1/3},
\eea
with $  \rho_p  = 3\times10^{14}\ {\rm g/cm}^3, T_R = 30$ MeV, and taking $ n = 5 $. The optical depth is given by 
\beq
\tau = \int_{r_x}^{\infty} \kappa \rho dr,
\eeq
where $ \kappa$ is the opacity.  To find the opacity, we start from the reduced mean Rosseland opacity 
\beq
\frac{1}{\kappa \rho} = \int_{m_x}^{\infty} \frac{15}{4 \pi^4 T^5} \frac{E_A^2 e^{E_A/T} \sqrt{E_A^2 - m_A^2}}{(e^{E_A/T}-1)^2} l_A dE_A,
\eeq
where $ l_A $ is the mean free path.

@@

\begin{figure}[th]
\includegraphics[width=9cm]{stages.eps}
\caption{The changes in the bounds generated as each stage of the calculation departs from Ref.~\cite{dent_etal12}. Interestingly the kinematics have apparently relatively little effect.}
\end{figure}

Let's put some text here.

\begin{figure}[th]
	\includegraphics[width=9cm]{endtoend.png}
	\caption{A comparison of the final result for the excluded region with the result from Ref.~\cite{dent_etal12}. The excluded region is clearly significantly less restrictive, and opens up the region of parameter space for $\epsilon \approx  1e-8-9 $.}
\end{figure}

@@

\appendix
\section{Bremsstrahlung diagrams}
	Each of the $p-p-\pi $ vertices contribute a factor of $g_{pp}$,  the pseudoscalar pion-nucleon coupling. $f_{pp}$, the pseudovector pion-nucleon coupling, is related by $g_{pp} = \frac{2M_N}{m_\pi} f_{pp} $. The $ p-p-A' $ vertex contributes $e\epsilon \gamma_\mu$, where $\epsilon$ is the mixing strength between the photon and dark photon. Labeling the contributions according to the Feynman diagrams in Fig. , the matrix elements are 
	\bea
	M_1 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_1} \bar{u}(p_3) \gamma_5 u(p_2) \bar{u}(p4) \gamma_5 (\slashed{p}_1 - q_a +M_N)\slashed{\varepsilon} u(p_1),\\
	M_2 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_1} \bar{u}(p_4) \gamma_5 u(p_2) \bar{u}(p3) \gamma_5 (\slashed{p}_1 - q_a +M_N)\slashed \varepsilon  u(p_1), \\
	M_3 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_2} \bar{u}(p_3) \gamma_5 u(p_1) \bar{u}(p4) \gamma_5 (\slashed{p}_2 - q_a +M_N)\slashed \varepsilon  u(p_2), \\
	M_4 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_2} \bar{u}(p_4) \gamma_5 u(p_1) \bar{u}(p3) \gamma_5 (\slashed{p}_2 - q_a +M_N)\slashed \varepsilon  u(p_2), \\
	M_5 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_3} \bar{u}(p_4) \gamma_5 u(p_1) \bar{u}(p3) \slashed{\varepsilon} (\slashed{p}_3 + q_a +M_N)\gamma_5 u(p_2),\\
	M_6 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_3} \bar{u}(p_3) \gamma_5 u(p_1) \bar{u}(p4) \slashed{\varepsilon} (\slashed{p}_3 + q_a +M_N)\gamma_5 u(p_2),\\
	M_7 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_4} \bar{u}(p_3) \gamma_5 u(p_2) \bar{u}(p4) \slashed{\varepsilon} (\slashed{p}_4 + q_a +M_N)\gamma_5 u(p_1),\\
	M_8 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_4} \bar{u}(p_4) \gamma_5 u(p_2) \bar{u}(p_3) \slashed{\varepsilon} (\slashed{p}_4 + q_a +M_N)\gamma_5 u(p_1),
	\eea
	where the exchange momenta are defined by $ k = p_2-p_4 $ and $ l = p_2-p_3 $, and the dark photon's polarization is given by $\varepsilon$. 
	For $pn$ scattering, four of the diagrams are simply the same as those given above. The only new Feynman diagram comes from bremsstrahlung off the internal pion, and results in  
	 \beq
	 M'_5 = \frac{4 M_N}{ m_\pi} \frac{f_{pn}^2 e \epsilon}{l^2-m_\pi^2
	 }  \frac{1}{(l-q_a)^2 - m_\pi^2} \bar{u}(p_4) \gamma_5 u(p_1) \bar{u}(p_3) u(p_2) (q_a - 2l)\cdot \varepsilon
	 \eeq
	 Note that these matrix elements agree with Ref.~\cite{dent_etal12}.
\acknowledgments

@@

\bibliography{draft}
